---
title: "DATA607 - HW 1"
author: "Michael Munguia"
date: "2/1/2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Overview
The article, [American Chess is Great Again](https://fivethirtyeight.com/features/american-chess-is-great-again/), deals with details how professional chess players transferred to play for the U.S. more than any other country in the years from 2000 to 2017. It also details some of the murkiness in the trade negotiations and (reportedly) large sums of money involved in these trades. Further, it also explores how the saturation of U.S. Federation players may tip the odds in the federation's favor in terms of establishing a world champion during the timeframe the article was written.


# Header
Each text block needs to have at least a header and non-header text...I guess just keep things organized.

# Loading the data
Here I load the data and examine the structure, which will help give me a sense of what is available and some of the first few values.
```{r, eval=TRUE}
chess <- read.csv("https://raw.githubusercontent.com/mijomu/DATA-607/master/HW1/transfers.csv", stringsAsFactors = FALSE)
str(chess)
```

# Renaming variables
Based on the above, I can see the variable could generally be renamed to make them more meaningful to anyone using the data downstream. Looking at the structure again shows this is more readily understandable.
``` {r, eval=TRUE}
colnames(chess) <- c("source_url", "player_id", "new_federation", "old_federation", "transfer_date")
str(chess)
```

# Reformatting dates
Imagining this might want to be extended into dates prior to 2000, we might want to reformat the date values. In this case, we know the existing data is from 2000-2017, so the modification is fairly easy. Passing the dataframe to the str function again shows us that we've not only gotten the values to look standardized, but we've also gone ahead and changed them over to a Date data type as opposed to text. This means they can now be used for calculations downstream.
``` {r, eval=TRUE}
str_date <- chess$transfer_date
str_len <- nchar(str_date)
actual_date <- paste0(substr(str_date, 1, str_len - 2), "20", substr(str_date, str_len - 1, str_len))

chess$transfer_date <- as.Date(actual_date, format = "%m/%d/%Y")
str(chess)
```

# Subsetting
The source URLs are not strictly necessary for any further analysis as per the article, so we will drop that variable. Should we need them again later on, we can obtain them through the previous dataframe and match against the player ID present on both dataframes.
```{r, eval=TRUE}
transfer_info <- subset(chess, select = -c(source_url))
str(transfer_info)
```

# Conclusions
To verify their conclusions, I would have to visit and explore the FIDE website further and either download more readily available data provided there or write a webscraper to pull any supporting data mentioned in the article. There was a good amount of financial information stated that was not available in the data set itself, which I would need to track down and incorporate into any verification. To extend the work, I think it would be interesting to find data on the various tournaments/championships along with their locations, monetary prizes and the winners to see if this U.S.' push to gather as many high level players has ultimately paid off.