---
title: "DATA 607 - Week 3 Assignment"
author: "Michael Munguia"
date: "February 11, 2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})
```

**Task 1:** Using the 173 majors listed in fivethirtyeight.com's [College Majors dataset](https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/), provide code that identifies the majors that contain either "DATA" or "STATISTICS"

```{r}
majors <- suppressMessages(read_csv(
  "https://raw.githubusercontent.com/fivethirtyeight/data/master/college-majors/majors-list.csv"))
(data_stat_majors <- majors %>% filter(str_detect(Major, "DATA|STATISTICS")))
```

**Task 2:** Write code that transforms the data below:

[1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"

[5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"  

[9] "elderberry"   "lime"         "lychee"       "mulberry"    

[13] "olive"        "salal berry"

Into a format like this:

c("bell pepper", "bilberry", "blackberry", "blood orange", "blueberry", "cantaloupe", "chili pepper", "cloudberry", "elderberry", "lime", "lychee", "mulberry", "olive", "salal berry")

The output above looks to be the results of a character vector printed out to a very narrow console. As such, the following would print out output that looks like the vector-style format desired. For readability, I store the character vector as `fruits_vegetables`:
```{r fruits, linewidth=90}
fruits_vegetables <- c("bell pepper", "bilberry", "blackberry", "blood orange", "blueberry",
                       "cantaloupe", "chili pepper", "cloudberry", "elderberry", "lime",
                       "lychee", "mulberry", "olive", "salal berry")

end_result <- paste0("c(\"", paste(fruits_vegetables, collapse = "\", \""), "\")")

writeLines(end_result)
```
The two exercises below are taken from R for Data Science, 14.3.5.1 in the on-line version:

**Task 3:** Describe, in words, what these expressions will match:

To make life a bit easier, when I use the phrase character/any character below, I mean any alpha-numeric character.  
  
* `(.)\1\1` - This regular expression matches a string containing the same three consecutive characters. It would match, for example, "abbbc" but not "abc".
* `"(.)(.)\\2\\1"` - This is a string representing a regular expression that matches a pair of any characters followed by the reverse order of the same pair. So, "abbbba" and "abba" are both viable matches.
* `(..)\1` - This regular expression matches any two characters followed by the same sequence of two characters. A match would be "abbbba" or "abab".
* `"(.).\\1.\\1"` - This is a string representing a regular expression matching a string with any character followed by any other character, followed by the initial character, a third character and the initial character again. Matches would include "abaca", "dedad".
* `"(.)(.)(.).*\\3\\2\\1"` - This is a string representing a regular expression that matches any three consecutive characters, followed by an optional fourth character, followed by the initial three characters in reverse order regardless of whether that fourth character was present. Matches are strings like "abcdcba" or "abccba".


**Task 4:** Construct regular expressions to match words that:

* Start and end with the same character. - `^(.).*\1$` or `"^(.).*\\1$"` as a string.
* Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.) - `*(.?...?).*\1"` or `".*(.?...?).*\\1"` as a string.
* Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.) - `(.?[A-Za-z].?){3,}` or `"(.?[A-Za-z].?){3,}"` as a string